Appendix 1: sdmTMB statistical model description

### 2022-03-23 doc/appendix-model.Rmd
X model matrices

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a spatial component (a random field), $\omega_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\s,t}$ represents a spatiotemporal component (a random field), $\epsilon_{\s,t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} \ldots \right)
\end{aligned}
$$

Within `sdmTMB()`, $\bm{X}_{1,\bm{s},t} \bm{\beta}$ is defined by the `formula` argument and represents the main-effect model matrix and a corresponding vector of coefficients. This main effect formula can contain optional penalized smoothers or non-linear functions as defined below.

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} \bm{X}_{2,s,t} + \ldots \right),\\
  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
\end{aligned}
$$

where $\gamma_t$ is an example optional time-varying regression parameter and $x_{\s,t}$ is the corresponding predictor value.
This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
For example, `time_varying ~ 0 + x`.
The first value is given an implicit uniform prior.
I.e., the same variable should not appear in the fixed effect formula since the initial value is estimated as part of the time-varying formula.
The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and, this case, the intercept should be omitted from the main effects (`formula ~ + 0 + ...` or `formula ~ -1 + ...`).

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \bm{X}_{3,t} \zeta_{\bm{s}} + \ldots \right),\\
  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
\end{aligned}
$$

where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant spatially for a given time $t$.
Currently, $\bm{X}_{3,t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.
For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.

>>>>

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}^{\mathrm{main}}$, $\bm{X}^{\mathrm{tvc}}$, and $\bm{X}^{\mathrm{svc}}$ represent design matrices (the superscript identifiers 'main' = main effects, 'tvc' = time varying coefficients, and 'svc' = spatially varying coefficients);
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a spatial component (a random field), $\omega_{\s} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\s,t}$ represents a spatiotemporal component (a random field), $\epsilon_{\s,t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} \ldots \right)
\end{aligned}
$$

Within `sdmTMB()`, $\bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta}$ is defined by the `formula` argument and represents the main-effect model matrix and a corresponding vector of coefficients. This main effect formula can contain optional penalized smoothers or non-linear functions as defined below.

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots +  \bm{X}^{\mathrm{tvc}}_{\s,t} \gamma_{t} + \ldots \right),\\
  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
\end{aligned}
$$

where $\gamma_t$ is an example optional time-varying regression parameter and $\bm{X}^{\mathrm{tvc}}_{\s,t}$ is the corresponding predictor value.
This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
For example, `time_varying ~ 0 + x`.
The first value is given an implicit uniform prior.
I.e., the same variable should not appear in the fixed effect formula since the initial value is estimated as part of the time-varying formula.
The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and, this case, the intercept should be omitted from the main effects (`formula ~ + 0 + ...` or `formula ~ -1 + ...`).

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \bm{X}^{\mathrm{svc}}_{\s, t} \zeta_{\bm{s}} + \ldots \right),\\
  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
\end{aligned}
$$

where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
Usually, $\bm{X}^{\mathrm{svc}}_{\s, t}$ would represent a predictor that is constant spatially for a given time $t$.
Currently, $\bm{X}^{\mathrm{svc}}_{t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.
For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.
